---
title: "h5"
author: "Mario Annau"
date: "21 May 2016"
output:
  ioslides_presentation:
    highlight: tango
institution: R/Finance 2016
subtitle: An Object Oriented Interface to HDF5
theme: metropolis
highlight: tango
vignette: >
  %\VignetteIndexEntry{RFin2016 Slides} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(h5)
library(zoo)
if (file.exists('test.h5')) file.remove('test.h5')
```

## Data Challenges

- Store large amounts of data, e.g. tick data 
- Retrieve subsets of data into memory
- Programming language independent
- Minimal setup requirements, single client
- High performance

## HDF5 Hierarchical Data Format

- Binary, language independent file format
- Developed by NCSA[^1] and the tri-labs, supported by NASA[^2]
- First released in 1998
- HDF5 building blocks are *Groups*, *Datasets* and *Attributes*
- *Datasets* are stored in *Groups*, similar to folders in file system
- Addition of metadata using *Attributes*

[^1]: National Center for Supercomputing Applications
[^2]: National Aeronautics and Space Administration

## Existing HDF5 Packages

- History of packages: **rhdf5** (**h5r**, **hdf5**, **activeH5**)
- Packages suffer in terms of usability and speed
- No active packages on CRAN

## h5: An Object Oriented Interface to HDF5

- Intuitive to use through (subset) operators
- **Rcpp** to interface library
- 200+ test cases with a coverage of more than 80%
- Available on CRAN and Github for all major platforms

## First Steps

```{r firststeps}
f <- h5file('test.h5')
g1 <- f['group1']
g1['mat'] <- matrix(1:9, nrow = 3)
g1['mat2'] <- matrix(11:19, nrow = 3)
h5attr(g1, 'attr1') <- 'This is Group 1'
f['group2/mat3'] <- matrix(21:29, nrow = 3)
sapply(list.datasets(f), function(x) f[x][, 1])
h5close(f)
```

## Extract Time Series Created from Pandas

**Python**:

```python
from pandas import date_range, DataFrame
from numpy import random
t = date_range('2010-01-01', '2016-01-01', freq='D').date
randmat = random.standard_normal((len(t), 3))
df = DataFrame(randmat, index=t)
df.to_hdf('ex-pandas.h5', 'testset')
```

**R**:
```{r pandas2, eval=FALSE}
f <- h5file('ex-pandas.h5', 'r')
dates <- as.Date(f['testset/axis1'][1:3], 
  origin = '0001-01-01') - 1
zoo(f['testset/block0_values'][1:3, ], order.by=dates)
```

## Conclusion and Outlook

- **h5** directly maps R data types and objects to HDF5
- Facilitates data exchange between languages like R, Python and Matlab.
- Available on CRAN for all major platforms
- Support for *data.frames* planned
- Project is open for contributors, pull requests!

\center https://github.com/mannau/h5


```{r cleanup, include=FALSE}
if (file.exists('test.h5')) file.remove('test.h5')
```
